---
title: "Debt Analysis: How Occupation Affects Creditworthiness" 
subtitle: "Exploratory Data Analysis (EDA) and Statistical Inference Project (STAS102)" 
author: "Mosia, Noor (229895735)"
date: "2024" 
date-format: "YYYY" 
title-block-banner: "#141c2b" 
title-block-banner-color: "#ffffff" 
title-block-style: manuscript 
code-line-numbers: true 
code-block-border-left: "#006b34" 
code-annotations: select 
code-block-bg: "#e5f0ea" 
code-fold: true
number-sections: true 
fontcolor: "#141c2b" 
format:    
  html: default   
  pdf: default 
css: style.css 
---

```{r}
#| label: setup_project  
#| message: false

knitr::opts_chunk$set(echo = TRUE, message = FALSE, cache = TRUE) 

# Load necessary libraries
library(tidyverse)
library(skimr)
library(confintr)

# read in data
my_data <- read_csv("credit_data.csv")

```

## Introduction

In this report we will be using a random sample of `r count(my_data)` individuals to look at the relationship between a debtor's occupation, number of late payments for debts, debt to income ratio and credit utilization ratio. Then analyse how these variables affect a debtor's credit rating. Furthermore, we will make use of statistical inference techniques to make decisions about the parameters related to debtors and their credit score based on the random sample we have acquired for this report.

The reason I chose this particular dataset is because I am interested in pursuing a career in the financial sector. This project will provide me with valuable foundational knowledge that is essential in understanding the industry.

## Data

### Variables Summary

-   **`occupation`**: Categorical variable representing the job or profession of the individuals in the random sample.

-   **`credit_rating`**: Categorical variable, It represents an individual's creditworthiness. It is determined by looking at an individual's history and repayment behaviors.

-   **`number_of_late_payments`**: Numerically discrete variable. Counts the number of times an individual made a late payment on their loan repayment.

-   **`debt_to_income_ratio`**: Continuous variable. Measures the ratio of an individual's monthly debt payments to their gross income.

-   **`credit utilization_ratio`**: Continuous variable. Represents the ratio of an individual's credit used and the total credit available to the them.

### Distribution Breakdown

The table below shows the distribution of the sample data terms of the occupations of each of the `r count(my_data)` individuals:

```{r}
#| label: distribution of occupations
occupationTable <-table(my_data$occupation)
occupationTable
```

### Preview of Sample Data

```{r}
#| label: datapreview 

# Display the first six entries
knitr::kable(head(my_data))

```

## Descriptive Statistics/ EDA

Exploratory Data Analysis (EDA) helps us understand the underlying patterns, trends, and relationships within the data. It involves summarizing the main characteristics of the data, often using visual methods.

### Summary Statistics

Summary statistics provide a quick overview of the data. They include measures of central tendency (mean, median) and dispersion (standard deviation, range).

```{r summarystats}

# using the skimr package, stating what I'd like to see on the table.
my_skim <- skim_with(
  numeric = sfl(mean = mean, 
                sd = sd, 
                min = min, 
                p25 = ~ quantile(., .25),
                med = median,
                p75 = ~ quantile(., .75),
                max = max),
  append = FALSE
)

# Summary statistics
my_skim(my_data)
```

### Variables in depth

The dataset includes 6 types of `occupation`: Clerk, engineer, manager, nurse, salesperson and teacher. Engineers and managers are known to be paid much more than clerks and salespeople. The range in the earnings allows us to have a good representation of the population based on the earnings, even those that are not included in the sample.

The categorical variable `credit_rating` indicates an individual's creditworthiness. It is used by credit providers to indicate the likelihood of an individual to pay back their debt. In our report, credit ratings are represented by 3 grades, poor, fair and good. poor would mean high risk of debtor not paying off their debt, fair means there is some risk but they are worth granting credit and good means the debtor has high creditworthiness and low risk of default.

`Number_of_late_payments` is a numeric discrete variable and it represents the number of payments made at a later date than the one agreed upon by the debtor and the credit provider. This is a good representation of a debtors creditworthiness and will provide us with good information about a debtor's reliability based on their occupation.

`Debt_to_income_ratio`, also refered to as DTI, is a metric that compares a debtor's total monthly debt payments to their gross monthly income. It is a good measure of how manageable a an individual's debt is. A high DTI affects a debtor's credit health in more than one way. Firstly, it indicates to credit providers that an individual is over-leveraged. Secondly, Any personal emergency, big or small, could result in a debtor missing their monthly payments. Understanding the DTI of different occupations can help identify occupations with high risk of financial distress.

`Credit_utilization_ratio` is the measure of credit a person is using divided by the amount of credit that is available to them. A low credit utilization ratio indicates good credit management by a debtor. Analyzing credit utilization ratios accross different occupations can help identify groups that are risk of high utilization, this info could help credit providers in creating financial education programs for these occupation groups.

### Distribution of occupations

Visualizing the distribution of `occupations` helps us understand their spread and central tendency. To begin, gives us an indication of the spread of the occupations in the dataset.

```{r}
#| label: fig-occupations
#| fig-cap: "Occupations Distribution"
#| fig-align: "center"

# Distribution of Occupations
ggplot(my_data, aes(x = occupation)) +
  geom_bar(fill = '#006B34') +
  labs(title = 'Distribution of Occupations', x = 'Occupations', y = 'Count')
```

The box plot above helps us understand the distribution better. From looking at the graph, we see that there is a good variation in occupations represented. The range of income between the occupations is good as we have clerks representing low income occupations, teachers and nurses representing medium income occupations and managers together with engineers representing high income occupations. The range income with allow us to use findings from this report to make decisions about other occupations that are not included in our sample data.

### Scatter plot of Debt to Income Ratio vs. Credit Utilization Ratio

This scatter plot will help us understand the relationship between our two continuous variables.

```{r}
#| label: fig-ratios
#| fig-cap: "Debt to Income Ratio  vs. Credit Utilization Ratio"
#| fig-align: "center"

# Scatter plot of Debt to Income Ratio vs. Credit Utilization Ratio
ggplot(my_data, aes(x = debt_to_income_ratio, y = credit_utilization_ratio)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(title = 'Debt to Income Ratio  vs. Credit Utilization Ratio', x = 'Debt to Income Ratio', y = 'Credit Utilization Ratio')
```

By inspecting the scatter plot above, we see a weak linear relationship between debt to income ratio and credit utilization ratio. We can also find the correlation coefficient by using the R built in function `cor()` to examine the relationship between the two variables. For these two variables we find the correlation coefficient to be `r cor(my_data$debt_to_income_ratio, my_data$credit_utilization_ratio)`, which supports the weak positive linear relationship that we see visually from the scatter plot. </br> What is being indicated by the red line is a linear regression model showing a general trend. The general trend is that as the debt to income ratio increases, the credit utilization ratio also increases slightly. This suggests that there are other factors that influence an individual's credit utilization ratio besides their debt to income ratio. Overall, the positive linear correlation provides some information, but it also suggests that further analyses using other variables needs to be done to understand the factors involved.

### Box plot of Debt to Income Ratio vs. Credit Utilization Ratio

```{r}
#| label: fig-boxplot
#| fig-cap: "Box plot of Debt to income ratio vs credit rating by occupation"
#| fig-align: "center"
# Boxplot of Total Monthly Production by Solar Panel Type and Battery Storage
ggplot(my_data, aes(x = occupation, y = debt_to_income_ratio, fill = as.factor(credit_rating))) +
  geom_boxplot() +
  labs(title = 'Debt to income ratio vs credit rating by occupation', x = 'Occupation', y = 'Debt to income ratio', fill = 'Credit Rating')

```

For `fair credit ratings`, engineers generally have higher quartiles. This shows that they tend to have a higher debt to income ratio even when maintaining a fair credit rating.

For `good credit ratings`, the median is fairly close between all occupations.

For `poor credit ratings`, there is an insignificant number of engineers and managers with a poor credit rating. Clerks have a significantly higher median debt to income ratio than all the other occupations.

## Statistical Inference

### Estimation

For this dataset, the discrete variable is `number_of_late_payments`. The poisson distribution is going to be fitted on this variable. By the method of moments process, the sample mean is used to estimate the unknown lambda parameter of the poisson distribution. So the sample mean of ``r mean(my_data$number_of_late_payments)`` will be used to estimate lambda.

```{r}
#| label: fig-poisson
#| fig-cap: "Plot of poisson distribution for number of late payments"
#| fig-align: "center"
options(scipen=999) # to see decimals not exponential notation

#Discrete variable
late_payments <- my_data$number_of_late_payments
payment_freq <- c(20, 57, 48, 23, 33, 13, 19,  9,  9,  5,  3,  3,  1,  1)
discrete_late_payments <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 14)

# Poison fit
lambda_hat <- mean(late_payments)
pois <- dpois(x=discrete_late_payments, lambda = lambda_hat)
pois[length(pois)] <- 1 - sum(pois[1:length(pois)-1])

ev_late_payments <- round(pois*nrow(my_data))
  
data <- data.frame(
  late_payments = as.factor(discrete_late_payments),
  observed = payment_freq,
  pois = ev_late_payments
)

#Data wrangling
data2 <- data |>
  pivot_longer(names_to = "shapes", values_to="values", observed:pois)

#plot
ggplot(data2, aes(x = late_payments, y= values, fill = shapes))+
  geom_bar(position="dodge", stat="identity")+
  labs(title="Plot of poisson distribution")
```

Looking at the graph, it is evident that the poisson distribution is not a good fit. We will try to further confirm this findings by performing a goodness-of-fit test.

### Goodness-of-fit Test

```{r}
options(warn=-1) # remove the warning message
fit <- chisq.test(x=payment_freq, p=pois)
fit
```

The goodness-of-fit-test confirms the visual finding of poisson ``r mean(my_data$number_of_late_payments)`` not being a suitable distribution to fit `number_of_late_payments`.

### Interval Estimation

```{r}
ci <- ci_mean(my_data$credit_utilization_ratio) 

# Preview
ci
```

The 95% confidence interval for credit utilization ratio is approximately \[0.3977805,0.4223015\]. This interval means the true value of the average credit utilization ratio is within this range. This information is important in determining the range of DTI in the population.

### Hypothesis testing

```{r}
dti_mean <- mean(my_data$debt_to_income_ratio)
```

Null Hypothesis (H0): The true mean for debt to income ratio is equal to ``r dti_mean``

Alternative Hypothesis (H1): The true mean for debt to income ratio is not equal to ``r dti_mean``

```{r}
t_test = t.test(x=my_data$debt_to_income_ratio, mu=dti_mean, alternative="two.sided")
t_test
```

The p-value from the t-test is equal to 1, indicating that we fail to reject the null hypothesis. Therefore, the true mean of the population for debt to income ration is ``r dti_mean``.

## Insights

Based on the exploratory data analysis and statistical inference, here are the key insights:

1.  **Distribution of Occupations**: The range in occupations, their salaries and skill level allow us to use info we get from analysing this distribution for other occupations not included in the dataset.

2.  **Occupation creditworthiness**: High earning jobs such as engineers and managers produce very few if any individuals with poor creditworthiness.

3.  **DTI and credit utilization ratio relationship**: Although there is some relationship between the two rations, DTI is affected by more than the credit utilization ratio.

4.  **DTI and credit utilization ratio relationship**: From the hypotheses test, we see that there is no significant difference between the average debt to income and the true mean of the population. This serves as a foundation for further research on the distribution, knowing that we are certain about the value of the true mean.


5.  **Distribution for discrete variable**: By doing a method of moments estimation and performing a goodness-of-fit test using the poisson distribution, we find that distribution is not a good fit for out discrete variable `number_of_late_payments`.

## Conclusion

This report provides an analysis on various various variables that can affect an individual's credit score. Through exploratory data analysis, we identified the distributions and relationships between variables. Statistical inference methods were used to draw conclusions about the data. Key insights highlighted the overall trends and relationships within the dataset. This analysis can help inform decisions related to solar power production and the factors that influence it.

The analyses suggest that a high paying occupation protects an individual from having a poor credit rating. It has also been confirmed that the true average for debt to income ratio is ``r dti_mean``.

## Recommendations

Based on the insights gained from this analysis, the following recommendations can be made:

1.  **Targeted Financial Literacy Programs**: Clerks have a high median debt to income ratio. It would be helpful to create tailor made education products for them in order to boost their credit health.

2.  **Try another estimation method**: Since we know that the poisson distribution is not a good fit, further analysis can be done using other distributions.

3.  **Using results from hypothesis test**: Since we have found the true mean for the population through a hypothesis test, we can use the results from this report to spot sampling errors in future analyses.











